---
title: R Code for allowing non binary random design matrix in the lme4 R package 
author:  Christophe Ambroise
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
---


# Mixed models and non  binary random design matrix

The few functions described below allow to consider 
non  binary random design matrix using the machinery of the 'lme4' R package.

Writing some classical mixed models for assessing the mixing abilitiy for cultivar mixture, we realized that no free R package was available (2017) for dealing with non binary random design matrix. Although 'lme4'is a widespread and effective package, it does only consider random design issued from group structures. 

In order to overcome this limitation we wrote a replacement function for dealing the input of the model and we relied  on the optimization machinery of the package for estimating the parameters. 

# Principle of the code
```{r eval=TRUE, include=TRUE}
library(lme4)
```

The core function  is a replacement for the function 'mkReTrms' of the  'lme4' package, which makes the random effect terms (by mainly creating the design matrix $Z$ and inpution the random covariance matrix of each factor). 

```{r mkReTrms-replacement, eval=TRUE, include=TRUE}
myReTrms<-function(ListZ,ListVar=NULL){
  reTrms<-list()
  reTrms$Zt    <- Matrix(t(Reduce('cbind',ListZ)),sparse=TRUE)
  reTrms$theta <- rep(1,length(ListZ))  # Initial Value of the covariance parameters
  reTrms$Lind  <- rep(1:length(ListZ),unlist(lapply(ListZ,ncol))) #an integer vector of indices determining the mapping of the elements of the theta vector to the "x" slot of Lambdat
  reTrms$Gp    <- as.integer(unname(cumsum(c(0,unlist(lapply(ListZ,ncol))))))
  reTrms$lower   <- rep(0,length(ListZ)) # lower bounds on the covariance parameters
  if (is.null(ListVar)) {reTrms$Lambdat <- Matrix(diag(rep(1,sum(unlist(lapply(ListZ,ncol))))),sparse=TRUE)} # transpose of the sparse relative covariance factor
  else {reTrms$Lambdat <- bdiag(lapply(ListVar,chol))}
   reTrms$Ztlist <- lapply(ListZ,function(Z) Matrix(t(Z),sparse=T))
  reTrms$cnms   <-  as.list(names(ListZ)) ; names(reTrms$cnms)<- names(ListZ)
  # Flist is Not very clean (to say the least... )
  reTrms$flist <- lapply(ListZ,function(Z) {flist.factor<- as.factor(colnames(Z)[apply(Z,1,function(x) which(rmultinom(n=1,size=1,prob =x)==1) )]);
                                            levels(flist.factor)<-colnames(Z); return(flist.factor)}) #NULL # list of grouping factors used in the random-effects terms (used for computing initial variance ??)
  return(reTrms)
}                
```


The main function is a replacement of the `lmer` function. Instead of specifying an R formula for describing the model,  the replacement  function takes as input 
  - the response vector
  - the fixed effect design matrix
  - the list of random effect design matrices
  - the corresponding list of covariance matrices


```{r lmer-replacement, eval=TRUE, include=TRUE}
mylmer <- function(Response, X, ListZ,ListVar=NULL, REML = TRUE){
  notNA<-!is.na(Response)       # Get rid of the NA
  Response<- Response[notNA]   # Find another solution about the NA ???
  X <- X[notNA,]
  ListZ <- lapply(ListZ,function(Z) Z<-Z[notNA,]) 
  fr<-model.frame(Response~.,data.frame(Response=Response,X) )               
  reTrms <- myReTrms(ListZ,ListVar)
  devfun <- mkLmerDevfun(fr, X, reTrms)
  opt <- optimizeLmer(devfun)
  return(mkMerMod(environment(devfun), opt, reTrms, fr = fr))
}
```

The extraction of the random effect needs also a modified function. 


```{r ranef-replacement, include=TRUE, eval=TRUE}
# My ranef replacement
myranef<-function(model,condVar=TRUE){
  re.cond.mode<-tapply(model@u,mymod@pp$Lind,function(x) x)
  names(re.cond.mode)<- names(model@cnms)

  if (condVar) {
    Zt<-model@pp$Zt
    D  <- sigma(model)* t(model@pp$Lambdat) %*% model@pp$Lambdat
    Sigma<- t(Zt)%*%  D %*%Zt + sigma(model)*diag(rep(1,ncol(Zt)))
    var.cond <- D - Zt %*%solve(Sigma) %*% t(Zt)
    var.cond <- diag(var.cond) 
    var.cond.mode <- tapply(var.cond,mymod@pp$Lind,function(x) x)
    }
  for (i in 1:length(re.cond.mode)) {
    re.cond.mode[[i]]<-data.frame(re.cond.mode[i])
    names(re.cond.mode[[i]])<-names(re.cond.mode[i])
    row.names(re.cond.mode[[i]]) <- levels(model@flist[[i]])
    
    if (condVar) attr(re.cond.mode[[i]],"postVar")=array(var.cond.mode[[i]],c(1,1,nrow(re.cond.mode[[i]])))
  }
  attr(re.cond.mode,"class") <- "ranef.mer" 
  re.cond.mode
}
```


# Example code
As specified above the model specification needs the following objects:
  - the response vector
  - the fixed effect design matrix
  - the list of random effect design matrices
  - the corresponding list of covariance matrices

If the last argument is empty each covariance matrix associated to the random effect will be spherical. 

### Estimation of two different model (with and without GMA)

```{r  example, eval=TRUE, include=TRUE}
load("cultivar-mix.Rdata")
ListZ <- list(GMA=Zg,SMA=Zc)
mymod<-mylmer(Response =  Phenotype[[1]],X, ListZ,REML=FALSE) 
mymod.gma<-mylmer(Response =  Phenotype[[1]],X, ListZ[1],REML=FALSE) 
```

Once the model is estimated classical outputs are available

```{r classical-outputs, eval=TRUE, include=TRUE}
summary(mymod)
sigma(mymod)
```

Models can be compared

```{r model-comparison}
anova(mymod.gma,mymod)
```


Alternative plots are available

```{r alternative-plots,fig.show=TRUE, include=TRUE}
library(lattice)
dotplot(myranef(mymod))
pr01 <- profile(mymod)
xyplot(pr01, aspect = 1.3)
confint(pr01,level=0.9)
splom(pr01)  
```

# Caution

The functions are still experimental and should be used at your own risk.
